---
title: "Week 3 discussion"
author: "Haylee Oyler"
format: html
editor: visual
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---


### Load libraries
```{r}
library(here)
library(tidyverse)
library(sf)
library(tmap)
```

## Read in data
```{r}
sb_protected_areas <- read_sf(here::here("data", "cpad_super_units_sb.shp")) %>% 
  st_transform("ESRI:102009")

sb_city_boundaries <- read_sf(here::here("data", "sb_city_boundaries_2003.shp")) %>%
  st_transform("ESRI:102009")

sb_county_boundary <- read_sf(here::here("data", "sb_county_boundary_2020.shp")) %>%
  st_transform("ESRI:102009")

aves <- read_sf(here::here("data", "aves_observations_2020_2024.shp")) %>%
  st_transform("ESRI:102009")
```

Find how many bird observations are within protected areas in Santa Barbara County
- Show the different outputs from a spatial subset and a spatial join
- Bonus Challenge: Try it out with a 5 km buffer around the protected areas too!
```{r}

```

Find the protected areas within 15 km of a city in Santa Barbara County
- Hint: Use dplyr::filter() to select a city from sb_city_boundaries
- Explore the different outputs with st_intersects(), st_intersection(), and st_within()
- Practice a distance-based join with st_is_within_distance()

Find the distance between your city of choice and a protected area of your choice
- Note: st_distance() finds the distance between the geometries’ edges
