---
title: "Week 3 discussion"
author: "Haylee Oyler"
format: html
editor: visual
execute: 
  warning: false
  message: false
editor_options: 
  chunk_output_type: console
---


### Load libraries
```{r}
library(here)
library(tidyverse)
library(sf)
library(tmap)
```

## Read in data
```{r}
sb_protected_areas <- read_sf(here::here("data", "cpad_super_units_sb.shp")) %>% 
  st_transform("ESRI:102009")

sb_city_boundaries <- read_sf(here::here("data", "sb_city_boundaries_2003.shp")) %>%
  st_transform("ESRI:102009")

sb_county_boundary <- read_sf(here::here("data", "sb_county_boundary_2020.shp")) %>%
  st_transform("ESRI:102009")

aves <- read_sf(here::here("data", "aves_observations_2020_2024.shp")) %>%
  st_transform("ESRI:102009")
```

# Task one

Find how many bird observations are within protected areas in Santa Barbara County
- Show the different outputs from a spatial subset and a spatial join
- Bonus Challenge: Try it out with a 5 km buffer around the protected areas too!
## Subsetting
```{r}
aves_pa_subset <- sb_protected_areas[aves, ]
nrows(aves_pa_subset)
```

```{r}
tm_shape(sb_protected_areas) +
  tm_polygons() +
  tm_shape(aves_pa_subset) +
  tm_dots()
```

## Joining
```{r}
aves_pa_join <- st_join(aves, sb_protected_areas) # keep all bird obs regardless of inside protected areas
nrow(aves_pa_join) # keeps all the original info from aves
# Keeps all our original data and associates new attributes to it. 
```

```{r}
tm_shape(sb_protected_areas) +
  tm_polygons() +
  tm_shape(aves_pa_join) +
  tm_dots()
```


## Buffer around PAs
```{r}
st_crs(sb_protected_areas)$units
# m
```

```{r}
PAs_5km <- st_buffer(sb_protected_areas, dist = 5000)
```

```{r}
aves_5km <- PAs_5km[aves, ]
nrow(aves_5km) #stretched the area of the polygons so it's larger 
```

# Task 2
Find the protected areas within 15 km of a city in Santa Barbara County
- Hint: Use dplyr::filter() to select a city from sb_city_boundaries
- Explore the different outputs with st_intersects(), st_intersection(), and st_within()
- Practice a distance-based join with st_is_within_distance()

```{r}
carp <- sb_city_boundaries %>% 
  filter(NAME == "Carpinteria")

pa_buffer <- st_buffer(carp, dist = 15000)

pa_in_buffer <- pa_buffer[sb_protected_areas, ]
```

```{r}
tm_shape(pa_buffer) +
  tm_borders() +
tm_shape(carp) +
  tm_polygons() +
tm_shape(pa_in_buffer) +
  tm_borders()
```


```{r}
pas_carp_intersect <- st_intersects(carp, 
                                    sb_protected_areas, 
                                    dist = units::set_units(15, "km"), 
                                    sparse = FALSE)

pas_within_carp_intersect <- sb_protected_areas[pas_carp_intersect, ]


pas_within_carp_log <- st_is_within_distance(carp, 
                                         sb_protected_areas, 
                                         
                                         sparse = FALSE)

pas_within_carp <- sb_protected_areas[pas_within_carp_log, ]
```


Find the distance between your city of choice and a protected area of your choice
- Note: st_distance() finds the distance between the geometries’ edges
